Extreme heterogeneity of influenza virus infection in single cells

### 2017-10-17 paper/paper.tex
eLife full submission

\begin{abstract}
Viral infection can dramatically alter a cell's transcriptome.
However, these changes have mostly been studied by bulk measurements on many cells.
Here we use single-cell mRNA sequencing to examine the transcriptional consequences of influenza virus infection.
We find extremely wide cell-to-cell variation in production of viral gene transcripts -- viral transcripts compose less than a percent of total mRNA in many infected cells, but a few cells derive over half their mRNA from virus.
Some infected cells fail to express at least one viral gene, and this gene absence partially explains variation in viral transcriptional load.
Despite variation in total viral load, the relative abundances of viral mRNAs are fairly consistent across infected cells.
Activation of innate immune pathways is rare, but some cellular genes co-vary in abundance with the amount of viral mRNA.
Overall, our results highlight the complexity of viral infection at the level of single cells.
\end{abstract}

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_manualedits.pdf}}
\caption{\label{fig:fluexpr}
Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
{\bf (A)} 
The fraction of viral mRNA from each viral gene for each cell. 
{\bf (B)}
Box plots showing the distribution of the fraction of viral mRNA per cell from each viral gene.
The black lines at the notches are the medians, and the tops and bottoms of boxes indicate the first and third quartiles.
Whiskers extend to the highest or lowest data point observed within 1.5x the interquartile range, outliers shown as circles.
Notches extend 1.58x the interquartile range divided by the square root of the number of observations. 
{\bf (C)}, {\bf (D)} 
The same plots, but only including cells for which we observed at least one molecule of each viral gene.
}
\figdata{The raw data are in \texttt{p\_flu\_expr.csv}.}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap_manualedit.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_sig\_cellular\_genes.csv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

@@

\begin{suppfile}
\caption{\label{suppfile:cellgenematrix}
The annotated cell-gene matrix in Matrix Market Format.}
\end{suppfile}

>>>>

\begin{abstract}
Viral infection can dramatically alter a cell's transcriptome.
However, these changes have mostly been studied by bulk measurements on many cells.
Here we use single-cell mRNA sequencing to examine the transcriptional consequences of influenza virus infection.
We find extremely wide cell-to-cell variation in the productivity of viral transcription -- viral transcripts comprise less than a percent of total mRNA in many infected cells, but a few cells derive over half their mRNA from virus.
Some infected cells fail to express at least one viral gene, but this gene absence only partially explains variation in viral transcriptional load.
Despite variation in viral load, the relative abundances of viral mRNAs are fairly consistent across infected cells.
Activation of innate immune pathways is rare, but some cellular genes co-vary in abundance with the amount of viral mRNA.
Overall, our results highlight the complexity of viral infection at the level of single cells.
\end{abstract}

@@

\begin{figure}
\centerline{\includegraphics[width=0.9\linewidth]{figures/p_flu_expr_manualedits.pdf}}
\caption{\label{fig:fluexpr}
Relative expression of influenza virus genes in highly infected cells (>5\% of total mRNA from virus).
{\bf (A)} 
The fraction of viral mRNA from each viral gene for each cell. 
{\bf (B)}
Box plots showing the distribution of the fraction of viral mRNA per cell from each viral gene.
The black lines at the notches are the medians, and the tops and bottoms of boxes indicate the first and third quartiles.
Whiskers extend to the highest or lowest data point observed within 1.5x the interquartile range, outliers shown as circles.
Notches extend 1.58x the interquartile range divided by the square root of the number of observations. 
{\bf (C)}, {\bf (D)} 
The same plots, but only including cells for which we observed at least one molecule of each viral gene.
}
\figdata{The raw data for all cells are in \texttt{p\_flu\_expr\_all.csv}.}
\figdata{The raw data for fully infected cells are in \texttt{p\_flu\_expr\_fullyinfected.csv}.}
\end{figure}

@@

\begin{figure}
\centerline{\includegraphics[width=0.7\linewidth]{figures/p_cellular_heatmap_manualedit.pdf}}
\caption{\label{fig:cellulargenes}
Cellular genes that co-vary in expression with the amount of viral mRNA in cells expressing all eight viral genes.
The columns are cells, ordered from left to right by the fraction of mRNA derived from virus.
Each row is a gene that is differentially expressed as a function of the fraction of mRNA derived from virus at a false discovery rate of 0.1. 
Genes for which the color goes from blue at left to red at right are expressed at higher levels in cells with more viral mRNA.
The scale bar indicates the number of standard deviations above or below the mean expression, truncated at 3-fold on both sides. 
}
\figdata{The full results of the differential expression test is in \texttt{p\_sig\_cellular\_genes.csv}.}
\figdata{The results of a gene-set analysis are in \texttt{p\_pathway\_enrichment.tsv}.}
\figsupp[Many genes that co-vary with viral load are involved in the oxidative stress response. \label{figsupp:ROStable}]{
Table delineating genes in Figure~\ref{fig:cellulargenes} that are associated with the response to oxidative stress~\citep{Duong:2017ju,Jung:2017jz,Lee:2017ba,Peuchant:2017hj,MacLeod:2016kt,Jiang:2016cf,Gorrini:2013jv,Miura:2013fr,Kim:2009gi,Banning:2005cz,Murray:2003ie,Doyle:1999ts}.
}{\includegraphics[width=0.7\linewidth]{figures/ROStable.pdf}}
\end{figure}

@@

\begin{suppfile}
\caption{\label{suppfile:cellgenematrix}
The annotated cell-gene matrix in Matrix Market Format.
\jdbcomment{This file is too large for the \textit{eLife} submission system. We will communicate with the editors to get it uploaded for a final accepted version, or we will post it on DataDryad. If you are a reviewer and need access, please contact the editor.}
}
\end{suppfile}
